Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-19 docs/proposals/NSF/ProjDesc/ProjDesc.tex
added travel doc and went through proposal

@@

% Specific
\newcommand{\uu}{\bvec{u}}
\newcommand{\xx}{\bvec{x}}
\newcommand {\qq} {\bvec{q}}
\newcommand{\nn}{{\mathbf{n}}}
\renewcommand{\ss}{{\mathbf{s}}}
\newcommand{\Vn}{V_\nn}
\newcommand{\CE}{C_E}
\newcommand {\ny}{n}
\newcommand {\bdry} {\partial B}
\newcommand{\Diff}{D}
\newcommand{\thL}{$\theta$--$L$}
\newcommand{\bd}{\partial}
\newcommand{\eeta}{\boldsymbol{\eta}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\ff}{\mathbf{f}}
\newcommand{\yy}{\mathbf{y}}

@@

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Ch’en Sinkhole in Belize. Bottom: examples of soil erosion. The bottom right shows the microscopic soil texture, exhibiting polydispersity and anisotropy.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

Only recently have these ideas been extended to the low-Reynolds-number regime that is applicable to porous media~\cite{mitchell2017generalized, Quaife2018}. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}[noitemsep]
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered.

@@

{\bf An ultimate goal of this proposal is to achieve high-fidelity simulations of $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. The color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

In addition to the work of the PIs, dynamic erosion of a porous media is a highly active subject as laboratory experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing represent rapid changes.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.

%----------------------------------------------------------------------%
\subsection{Theoretical Foundation and Governing Equations}

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations \begin{equation}
\label{eqn:StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the
fluid viscosity, and $\Omega$ is the fluid domain. Linear elliptic PDEs,
such as the Stokes equations, can be reformulated as a boundary integral
equation (BIE).  BIEs have several numerical advantages including a
dimension reduction, automatic mass conservation, and high-order
accuracy. An equivalent BIE formulation of the Stokes equations in
Eq.~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

@@

Whilst the most immediate proposed work on computational methods to
simulate erosion will utilize~\eqref{eqn:StokesEq}, the later stages of this project will also incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

@@

\paragraph{Broader Impacts:} The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The high-fidelity, efficient computational methods proposed here will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

In addition to these societal impacts, the work proposed here will spur the training of mathematicians and computational scientists in fundamentally cross-disciplinary research of great societal importance. Further, the PIs will engage in educational and outreach activities to impact future generations of mathematical scientists, with particular focus on women and underrepresented minorities (URM). Both PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in undergraduate, graduate, and postdoctoral research.

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed research on porous-media erosion and dispersive transport. Large ensembles of the high-fidelity simulations will inform the coarse-grained models to be developed in the later stages of the proposed work.
  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Scientific Computing, Mathematics, and others.
  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close interaction and cooperation between them.
\end{itemize}

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion \cite{chi-moo-qua2019}.

\paragraph{Outline of the Proposal:}

\todo[inline]{Summary of what is in the proposal.}

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
The goal of the proposed research is to develop new numerical methods to
simulate erosion processes. Specific research tasks will include:
\begin{itemize}[noitemsep]
  \item Simulating chemical erosion.

  \item Characterizing anisotropy and permeability to inform
    homogenized models such as Darcy's relation.

  \item Simulating the transport of finite Peclet number concentrations
    inside eroded porous geometries.

@@

  \item Developing new models to characterize the effect of erosion on
    pore size dynamics.

  \item Developing new parameterizations using modern machine learning
    techniques.

  \item Validating simulations with in-house experimental results.

  \item \todo[inline]{Something about transport, sedimentation, etc.}
\end{itemize}

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $C$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{C}{t} + \uu \cdot \grad C = \Diff \Lap C.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{C}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $\mathcal{O}(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables more densely packed configurations. The color in Fig~\ref{80circ8vort} represents the vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity.  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. As erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude as in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows how certain medium properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies disintegrate. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially. The much more modest decrease of area fraction is not sufficient to account for this exponential rate. Hence the reshaping process, in particular the formation of channels, plays a pivotal role the medium's macroscopic resistivity to flow.

Lastly, the right-most plot of Fig.~\ref{fig3} shows new, preliminary results on how medium anisotropy evolves during the reshaping process of erosion. Anisotropy calculated as the ratio of vertical to horizontal resistivity, with the vertical resistivity computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

>>>>

%\captionsetup{belowskip=-20pt}

@@

% Specific
\newcommand{\uu}{\bvec{u}}
\newcommand{\xx}{\bvec{x}}
\newcommand {\qq} {\bvec{q}}
\newcommand{\nn}{{\mathbf{n}}}
\renewcommand{\ss}{{\mathbf{s}}}
\newcommand{\Vn}{V_\nn}
\newcommand{\CE}{C_E}
\newcommand {\ny}{\mathbf{n}}
\newcommand {\bdry} {\partial B}
\newcommand{\Diff}{D}
\newcommand{\thL}{$\theta$--$L$}
\newcommand{\bd}{\partial}
\newcommand{\eeta}{\boldsymbol{\eta}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\ff}{\mathbf{f}}
\newcommand{\yy}{\mathbf{y}}

@@

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep,topsep=0pt]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} \em Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Ch’en Sinkhole in Belize. Bottom: examples of soil erosion. The bottom right shows the microscopic soil texture, exhibiting polydispersity and anisotropy.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

Only recently have these ideas been extended to the low-Reynolds-number regime that is applicable to porous media~\cite{mitchell2017generalized, Quaife2018}. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}[noitemsep,topsep=0pt]
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered.

@@

{\bf An ultimate goal of this proposal is to simulate $\boldsymbol{\mathcal{O}(10^2-10^3)}$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with high-order accuracy and the computational efficiency required to run large ensembles for statistical analysis.} A major step towards this goal is illustrated in Fig.~\ref{80circ8vel}, which features high-fidelity erosion simulations of an initial configuration of 80 bodies of high volume fraction $50 \%$. Color indicates the velocity magnitude, which highlights the spontaneous formation of channels during the erosion process. These channels arise from small variations in the initial condition. Once formed, the larger channels transmit more flow which reinforces their growth. The process of channel formation can dramatically alter medium properties and create high levels of anisotropy, which directly impacts dispersion of transported quantities such as contaminants.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vel.pdf}
\caption{\label{80circ8vel} \em High-fidelity numerical simulations of a configuration of densely-packed solid bodies undergoing mechanical erosion due to an intervening Stokes flow. The color indicates velocity magnitude which highlights the channel formation process. The solid bodies develop flat faces and sharp corners during evolution. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

In addition to the work of the PIs, dynamic erosion of a porous media is a highly active subject as laboratory experiments on erosion of colloidal systems are currently being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments are closely related to our proposed computational study, thus offering a great opportunity for comparison. Furthermore, {\bf we propose performing a set of experiments conducted at the nearby Geophysical Fluid Dynamics Institute (see Section~\ref{sec:experiments}), which would help to guide, calibrate, and validate the computations (see letter of collaboration).}

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse or hydraulic fracturing represent rapid changes.  Most likely, all of the mechanisms listed above---erosion, dissolution, transport, compaction, and gravitational collapse---contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity.  Sinkhole formation has been explored in preliminary experiments conducted at the Geophysical Fluid Dynamics Institute at Florida State University~\cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of the proposed project is to merge these two lines of inquiry to gain a deeper understanding of the influence of mechanically or chemically induced erosion on gravitational collapse.  Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase} to quantitatively describe and predict sinkhole collapse.  
\todo[inline]{The last paragraph of this sections seems out of place.  Maybe it just needs to be reversed with the one before it?}
%----------------------------------------------------------------------%
\subsection{Theoretical Foundation and Governing Equations}

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:StokesEq}
  \mu \Delta \uu = \grad p, \quad
  \grad \cdot \uu = 0, \quad \xx \in \Omega,
\end{align}
%\begin{equation}
%\label{eqn:StokesEq}
%\begin{split}
%\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
%\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
%\end{split}
%\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the fluid viscosity, and $\Omega$ the fluid domain. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

@@

Whilst the most immediate proposed work on computational methods to simulate erosion will utilize~\eqref{eqn:StokesEq}, the later stages of this project will incorporate~\eqref{DarcyEq}, or closely related variants known as multiphase models \cite{breward2003multiphase, cogan2010multiphase, Imma2019, eastham2019multiphase}, in order to develop coarse-grained descriptions for porous medium evolution and the associated dispersion within. These models will rely on statistical analysis of the data produced by the high-fidelity erosion simulations in order to parameterize the effects of erosion, leading to deterministic or stochastic reduced-order models for porous-medium evolution. 
%and erosion, combined with other external effects such as gravitationally-induced dynamics.

@@

\paragraph{Broader Impacts:} The proposed research could lead to a range of societal benefits, including better management of water resources, accurate prediction of contaminant transport, and an understanding of the long-term effects of human activities, such as hydraulic fracturing or groundwater pumping, on porous media and water resources. In such scenarios, the slow timescales of mechanical and chemical erosion are coupled to the faster timescales of human-induced changes and gravitational collapse. Currently, the relationship between these scales is poorly understood. The proposed high-fidelity, efficient computational methods will enable a deeper understanding of these processes, which is the first step towards developing effective strategies for managing water resources and mitigating catastrophic events such as contamination or structural collapse. For example, the insights gained here could potentially help identify locations in natural aquifers vulnerable to contamination and/or collapse.

In addition to these societal impacts, the proposed work will spur the training of mathematicians and computational scientists in fundamentally cross-disciplinary research of great societal importance. Further, the PIs will engage in educational and outreach activities to impact future generations of mathematical scientists, with particular focus on women and underrepresented minorities (URM). Both PIs are already actively involved in graduate and undergraduate research, as well as outreach to K-12 students. Further, we have successfully recruited women and underrepresented minorities in high school, undergraduate, graduate, and postdoctoral research.

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed research on porous-media erosion and dispersive transport. Large ensembles of the high-fidelity simulations will inform the coarse-grained models to be developed in the later stages of the proposed work.

  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019, cur-spe-hie-obr-goo-qua2018} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Mechanical Engineering, Scientific Computing, Mathematics, and others.

  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close interaction and cooperation between them.
\end{itemize}

@@

The PIs co-authored a paper in the {\em Journal of Computational Physics} describing numerical methods for simulating erosion in the Stokes flow regime~\cite{Quaife2018}, the regime relevant for the currently proposed groundwater applications. Very recently the PIs, along with postdoctoral scholar Shang-Huan Chiu, submitted a second paper on newly developed methods for precise characterization of transport and dispersion within the special class of porous-media configurations formed by erosion~\cite{chi-moo-qua2019}.

%----------------------------------------------------------------------%
\section{Proposed Research} 
%----------------------------------------------------------------------%
The goal of the proposed research is to develop computational methods to better understand erosion processes. Specific research tasks will include:
\begin{enumerate}[noitemsep,label=(\arabic*)]
  \item Simulating chemical erosion, and transport, including sedimentation and compaction, of eroding bodies

  \item Characterizing anisotropy, permeability, and pore size dynamics to inform homogenized models such as Darcy's relation~\eqref{DarcyEq}.

  \item Simulating the transport of finite Peclet number concentrations in eroded porous geometries.

@@

  \item Developing and validating new parameterizations using machine learning techniques, numerical simulations, and in-house experiments.

\end{enumerate}

@@

The development of accurate and efficient numerical methods to solve the Stokes equations enables high-fidelity simulations of the erosion process. We will consider erosion induced by either fluid-mechanical forces or by chemical reaction, i.e.~dissolution. In the former case, which we have already begun to investigate~\cite{Quaife2018}, erosion result from the shear stress $\tau = -\left(\nabla \uu + \nabla \uu^T \right)\nn \cdot \ss$, where $\nn$ and $\ss$ are the unit normal and tangent vectors on the surface respectively and $\uu$ is the velocity field as computed from the incompressible Stokes equations~\eqref{eqn:StokesEq}. The motion of the interface is proportional to the absolute value of shear 
\begin{align}
  \Vn = \CE \, \abs{\tau}.
\end{align}
In the case of chemically-induced erosion, a solute concentration field $c$ undergoes advection and diffusion:
\begin{align}
  \label{Ceq}
  \pd{c}{t} + \uu \cdot \grad c = \Diff \Lap c.
\end{align}
Material removal is then governed by Fick's law with interface velocity
\begin{equation}
\label{DissVn}
\Vn = \Diff \pd{c}{\ny}.
\end{equation}
Both erosion processes have been investigated by the PIs and collaborators in the context of single bodies in high Reynolds-number flows~\cite{Ristroph2012, MoorePOF2013, HuangJFM2015, MooreCPAM2017}. In the context of a porous media at low Reynolds-number flows, numerical methods have been developed by the PIs to simulate shear-induced erosion~\cite{Quaife2018, chi-moo-qua2019}. These simulations combine highly accurate boundary integral methods with stable interface evolution techniques.  The boundary integral solver is accelerated by the fast multipole method~\cite{gre-rok1987, gre-gre-may1992} to achieve optimal $\mathcal{O}(N)$ complexity. Meanwhile, the so-called {\thL} method is used for interface evolution to prevent distortion and tangling of the interface meshes~\cite{hou1994removing, MoorePOF2013, Quaife2018}.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} \em Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

Figs.~\ref{80circ8vel} and~\ref{80circ8vort} show preliminary results of 80 bodies, initially of random position and size, undergoing mechanical erosion in Stokes flow. This simulation extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed barycentric quadrature scheme~\cite{bar2014, bar-wu-vee2015, chi-moo-qua2019}, which enables more densely packed configurations. The color in Fig.~\ref{80circ8vort} represents the vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity.  Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. As erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. This channelization process is made even more apparent by viewing the velocity magnitude in Fig.~\ref{80circ8vel}.

The creation and reinforcement of channels significantly impacts the macroscopic properties of the porous medium. Fig.~\ref{fig3} shows how certain medium properties change in time. First, the left-most plot shows the rate at which the area fraction decreases as the bodies disintegrate. Notice that the rate of area reduction does not change significantly during the simulation. The middle plot shows the resistivity, or inverse permeability of the medium. In the simulations, resistivity is calculated by measuring the total flux and the pressure upstream and downstream, then using Darcy's relation~\eqref{DarcyEq} to infer the horizontal permeability $k_x$. Naturally, as bodies erode, the resistivity they provide decreases in time. Note that, unlike the area plot, the vertical axis is logarithmic and so the straight line observed at early times indicates that the resistivity decreases exponentially.  The much more modest decrease of area fraction is not sufficient to account for this exponential rate. Hence the reshaping process, in particular the formation of channels, plays a pivotal role in the medium's macroscopic resistivity to flow.

Lastly, the right-most plot of Fig.~\ref{fig3} shows new, preliminary results on how medium anisotropy evolves during the reshaping process of erosion. Anisotropy is calculated as the ratio of vertical to horizontal resistivity, with the vertical resistivity computed by rotating the configuration of bodies by 90 degrees. This plot shows the anisotropy to initial increase with time. Thus, the initially isotropic state becomes erased as bodies streamline and form primarily horizontal channels. Remarkably, the medium reaches an anisotropy of over 10, meaning that erosion can lead to a configuration that resists flow 10 times more in the vertical direction than the horizontal direction.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} \em Left: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$.  Middle: The resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.} \end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
